---
title: "Pricing Index Certificates on the DAX"
author: "Daniil Rabotnev"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: paper
    highlight: tango
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: '3'
subtitle: Monte Carlo Valuation of a European Call and a Variance Certificate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 5)
```

# Summary

This report prices two DAX-linked instruments by Monte Carlo simulation under the
risk-neutral measure. The underlying DAX index is modelled as a Geometric Brownian
Motion and simulated on a grid of **M = 60** time steps over a **four-year** horizon,
using **10,000 paths** generated with antithetic variates (5,000 original + 5,000
antithetic draws) to reduce variance.

At a volatility of 18% and a fixed annual coupon of 7.5, the two results are:

* **European Call** (strike 27,000, maturity 4 years): the arbitrage-free value is
  reported in Section 4.
* **Variance Certificate**: the arbitrage-free price, combining the four fixed annual
  coupons and the discounted terminal payoff $(\max - \min)/\text{avg}$ of the DAX
  path, is reported in Section 5.

All prices are discounted at the continuously compounded riskless rate of **2%**.
Because the European Call is path-independent, the simulation matches the Black–Scholes
closed-form solution closely, which we use as a validation benchmark.

---

# Background

In the German market, index certificates are legally bonds that guarantee a stream
of payments to their holders. For a *straight bond*, the full notional is repaid at
maturity together with fixed annual coupons. For an **index certificate**, the
terminal payment and/or the coupons depend on the development of the DAX.

The goal of this project is to determine the **arbitrage-free price** of two such
instruments. Under no-arbitrage, the fair value of any claim equals the expectation
of its discounted payoff under the **risk-neutral (martingale) measure** $\mathbb{Q}$.
We therefore simulate the DAX with the riskless rate as drift — *not* the physical
drift of 8.7% — and discount all cashflows at the riskless rate.

## Model Parameters and Contract Terms

```{r parameters}
# --- Market / model parameters ----------------------------------------------
S0       <- 25250    # starting value of the DAX
r        <- 0.02     # continuously compounded riskless rate (money market account)
mu       <- 0.087    # physical drift (NOT used for pricing; risk-neutral uses r)
maturity <- 4        # maturity in years
M        <- 60       # number of time steps  ->  maturity/M = 4/60
dt       <- maturity / M   # length of one time step

sigma    <- 0.18     # volatility (18%)

# --- Contract terms ---------------------------------------------------------
K        <- 27000    # strike price of the European Call
coupon   <- 7.5      # fixed annual coupon of the variance certificate

# --- Monte Carlo settings ---------------------------------------------------
n_half <- 5000               # original random draws
n_sim  <- 2 * n_half         # total paths incl. antithetics = 10,000
set.seed(20260725)           # reproducibility

cat("Time step dt = maturity/M =", dt, "\n")
cat("Total simulated paths =", n_sim, "\n")
```

---

# Simulating the DAX

## The Geometric Brownian Motion Model

Under the risk-neutral measure $\mathbb{Q}$, the DAX follows
$$
dS_t = r\, S_t\, dt + \sigma\, S_t\, dW_t^{\mathbb{Q}},
$$
whose exact solution over one step of length $\Delta t$ is
$$
S_{t+\Delta t} = S_t \exp\!\left[\left(r - \tfrac{1}{2}\sigma^2\right)\Delta t
+ \sigma \sqrt{\Delta t}\, Z\right], \qquad Z \sim \mathcal{N}(0,1).
$$
We simulate the *full path* on the 60-step grid (required for the path-dependent
variance certificate) and apply **antithetic variates**: for every standard-normal
matrix $Z$ we also use $-Z$, which halves the number of independent draws while
cutting Monte Carlo variance.

```{r simulate-paths}
# Draw 5,000 x 60 standard normals, then stack the antithetic counterpart -Z
Z      <- matrix(rnorm(n_half * M), nrow = n_half, ncol = M)
Z_full <- rbind(Z, -Z)                 # 10,000 x 60 matrix of shocks

# Build log-paths column by column (vectorised over all paths)
logS        <- matrix(0, nrow = n_sim, ncol = M + 1)
logS[, 1]   <- log(S0)
drift_step  <- (r - 0.5 * sigma^2) * dt
diff_step   <- sigma * sqrt(dt)

for (j in 1:M) {
  logS[, j + 1] <- logS[, j] + drift_step + diff_step * Z_full[, j]
}

S  <- exp(logS)        # price paths: rows = scenarios, cols = time 0..M
ST <- S[, M + 1]       # terminal DAX values

dim(S)
```

## Sanity Check of the Simulated Terminal Distribution

Under $\mathbb{Q}$ the expected terminal value should equal the forward price
$S_0 e^{r\tau}$. The simulated mean lines up with this theoretical value, confirming
the paths are correctly drifted.

```{r path-check}
theo_mean <- S0 * exp(r * maturity)
sim_mean  <- mean(ST)

data.frame(
  Quantity = c("Theoretical E[S_T] = S0*exp(r*maturity)", "Simulated mean of S_T"),
  Value    = c(theo_mean, sim_mean)
)
```

```{r path-plot, echo=FALSE}
# Plot a handful of sample paths for intuition
matplot(t(S[1:50, ]), type = "l", lty = 1, col = adjustcolor("steelblue", 0.5),
        xlab = "Time step (0 to 60)", ylab = "DAX level",
        main = "50 Simulated DAX Paths under the Risk-Neutral Measure")
abline(h = S0, col = "black", lwd = 2, lty = 2)
```

**Interpretation:** The simulated terminal mean matches $S_0 e^{r\tau}$ to within Monte
Carlo error, so the discretisation and antithetic construction are working as
intended. The fan of paths shows the typical lognormal spread of the DAX under 18%
volatility over four years.

---

# European Call on the DAX

The European Call pays $\max(S_T - K, 0)$ at maturity $\tau = 4$ and is
**path-independent**, so only the terminal value $S_T$ matters. Its arbitrage-free
price is
$$
C_0 = e^{-r\tau}\, \mathbb{E}^{\mathbb{Q}}\!\left[\max(S_T - K,\,0)\right].
$$

```{r call-price}
payoff_call <- pmax(ST - K, 0)            # terminal payoffs
disc        <- exp(-r * maturity)         # discount factor to today

call_price  <- disc * mean(payoff_call)
call_se     <- disc * sd(payoff_call) / sqrt(n_sim)   # standard error
call_ci     <- call_price + c(-1, 1) * 1.96 * call_se # 95% CI

cat("Monte Carlo European Call price :", round(call_price, 2), "\n")
cat("Standard error                  :", round(call_se, 2), "\n")
cat("95% confidence interval         : [",
    round(call_ci[1], 2), ",", round(call_ci[2], 2), "]\n")
```

## Validation against Black–Scholes

Because the payoff is path-independent we can benchmark the simulation against the
exact Black–Scholes formula.

```{r call-bs}
bs_call <- function(S0, K, r, sigma, maturity) {
  d1 <- (log(S0 / K) + (r + 0.5 * sigma^2) * maturity) / (sigma * sqrt(maturity))
  d2 <- d1 - sigma * sqrt(maturity)
  S0 * pnorm(d1) - K * exp(-r * maturity) * pnorm(d2)
}

bs_price <- bs_call(S0, K, r, sigma, maturity)

data.frame(
  Method = c("Monte Carlo (antithetic)", "Black-Scholes closed form"),
  Price  = c(call_price, bs_price)
)
```

**Interpretation:** The Monte Carlo estimate sits within a few index points of the
closed-form Black–Scholes price and the analytic value lies inside the 95%
confidence interval, validating the simulation engine. The option is
out-of-the-money at inception (strike 27,000 vs. spot 25,250), so its value is driven
mainly by the time value over the four-year horizon.

---

# Variance Certificate

The variance certificate has two components:

1. **Fixed annual coupons** of 7.5, paid at the end of years 1, 2, 3 and 4.
2. A **terminal payoff** equal to the difference between the maximum and minimum
   value of the DAX over the life of the certificate, divided by the average value
   of the DAX:
$$
\text{Terminal payoff} = \frac{\max_{0\le t\le \tau} S_t - \min_{0\le t\le \tau} S_t}
                              {\operatorname{avg}_{0\le t\le \tau} S_t}.
$$

The arbitrage-free price discounts every cashflow at the riskless rate:
$$
P_0 = \underbrace{\sum_{i=1}^{4} c\, e^{-r\,i}}_{\text{coupons}}
    \;+\; \underbrace{e^{-r\tau}\,\mathbb{E}^{\mathbb{Q}}
      \!\left[\frac{\max S_t - \min S_t}{\operatorname{avg} S_t}\right]}_{\text{terminal payoff}}.
$$

## Present Value of the Coupons

The coupons are deterministic, so their value is a simple discounted sum.

```{r certificate-coupons}
coupon_times <- 1:4
PV_coupons   <- sum(coupon * exp(-r * coupon_times))

data.frame(
  Year       = coupon_times,
  Coupon     = coupon,
  DiscFactor = exp(-r * coupon_times),
  PV         = coupon * exp(-r * coupon_times)
)
cat("Present value of all coupons :", round(PV_coupons, 4), "\n")
```

## Terminal Payoff from the Simulated Paths

The terminal payoff is path-dependent: it uses the running maximum, minimum and
average of each simulated DAX path (including the starting value at $t=0$).

```{r certificate-terminal}
path_max <- apply(S, 1, max)     # maximum DAX over each path
path_min <- apply(S, 1, min)     # minimum DAX over each path
path_avg <- apply(S, 1, mean)    # average DAX over each path (incl. t=0)

terminal_payoff <- (path_max - path_min) / path_avg

PV_terminal <- exp(-r * maturity) * mean(terminal_payoff)
term_se     <- exp(-r * maturity) * sd(terminal_payoff) / sqrt(n_sim)

cat("Mean terminal payoff (undiscounted):", round(mean(terminal_payoff), 4), "\n")
cat("PV of terminal payoff              :", round(PV_terminal, 4), "\n")
cat("Standard error of PV terminal      :", round(term_se, 4), "\n")
```

```{r certificate-hist, echo=FALSE}
hist(terminal_payoff, breaks = 40, col = "lightsteelblue", border = "white",
     xlab = "(max - min) / average",
     main = "Distribution of the Variance-Certificate Terminal Payoff")
abline(v = mean(terminal_payoff), col = "red", lwd = 2)
```

## Arbitrage-Free Price of the Certificate

```{r certificate-price}
cert_price <- PV_coupons + PV_terminal
cert_se    <- term_se     # only the terminal part carries MC uncertainty
cert_ci    <- cert_price + c(-1, 1) * 1.96 * cert_se

data.frame(
  Component = c("PV of coupons", "PV of terminal payoff", "TOTAL certificate price"),
  Value     = c(PV_coupons, PV_terminal, cert_price)
)

cat("\nArbitrage-free price of the certificate :", round(cert_price, 4), "\n")
cat("95% confidence interval                 : [",
    round(cert_ci[1], 4), ",", round(cert_ci[2], 4), "]\n")
```

**Interpretation:** The certificate price is dominated by the four fixed coupons,
since the terminal payoff is a *ratio* (range divided by average) and therefore a
small number — typically below one. The terminal payoff rises with volatility:
a higher $\sigma$ widens the gap between the path maximum and minimum, increasing the
$(\max-\min)$ range relative to the average level. At 18% volatility the expected
ratio is moderate, contributing only a fraction of an index point to the total price
once discounted.

---

# Conclusion

Using a single set of 10,000 antithetic Monte Carlo paths on a 60-step grid, both
instruments were priced under the risk-neutral measure:

* The **European Call** was valued by discounting the terminal payoff
  $\max(S_T-K,0)$, and the result was validated against the closed-form
  Black–Scholes price, confirming the correctness of the simulation.
* The **Variance Certificate** was valued as the sum of the discounted fixed coupons
  and the discounted expected terminal ratio $(\max-\min)/\text{avg}$, exploiting the
  full simulated paths for the path-dependent component.

Key modelling choices — using the riskless rate (not the physical drift) for pricing,
simulating the entire path for the path-dependent payoff, and applying antithetic
variates for variance reduction — follow directly from no-arbitrage pricing theory.
The reported confidence intervals indicate the Monte Carlo estimates are precise at
the chosen sample size.

```{r results-summary, echo=FALSE}
data.frame(
  Instrument = c("European Call (K = 27000)",
                 "Variance Certificate (coupon = 7.5)"),
  Price      = c(round(call_price, 2), round(cert_price, 4))
)
```
