---
title: "Pricing Index Certificates on the DAX"
author: "Daniil Rabotnev"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: paper
    highlight: tango
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: '3'
subtitle: Monte Carlo Valuation of a European Call and a Variance Certificate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 5)
devtools::load_all("..")
```

# Summary

This report prices two DAX-linked instruments by Monte Carlo simulation under
the risk-neutral measure, using the `daxcerts` functions in `R/` — the same
code the companion reports analyse and the test suite covers.

At a volatility of 18% and a fixed annual coupon of 7.5:

* **European Call** (strike 27,000, maturity 4 years) — Section 4, validated
  against the Black–Scholes closed form.
* **Variance Certificate** — Section 5, the four fixed annual coupons plus the
  discounted terminal payoff $(\max - \min)/\text{avg}$ of the DAX path.

All cashflows are discounted at the continuously compounded riskless rate of
**2%**.

Three results from the companion reports shape how the numbers below are
produced and how much weight they carry:

* Antithetic sampling is used for the call and **not** for the certificate. It
  reduces variance for the first and increases it for the second — see
  `02-variance-reduction.Rmd`.
* Standard errors are computed on antithetic *pairs* rather than pooled draws.
  The pooled formula misstates precision by roughly a quarter.
* The certificate price on a 60-step grid is biased downwards by about 10%
  relative to continuous monitoring — see `03-discretisation.Rmd`. The value
  quoted here inherits that bias and should be read as a lower bound.

---

# Background

In the German market, index certificates are legally bonds that guarantee a
stream of payments to their holders. For a *straight bond*, the full notional is
repaid at maturity together with fixed annual coupons. For an **index
certificate**, the terminal payment and/or the coupons depend on the development
of the DAX.

The goal is to determine the **arbitrage-free price** of two such instruments.
Under no-arbitrage, the fair value of any claim equals the expectation of its
discounted payoff under the **risk-neutral (martingale) measure** $\mathbb{Q}$.
The DAX is therefore simulated with the riskless rate as drift — *not* the
physical drift of 8.7% — and all cashflows are discounted at the riskless rate.

## Model Parameters and Contract Terms

```{r parameters}
S0       <- 25250    # starting value of the DAX
r        <- 0.02     # continuously compounded riskless rate
mu       <- 0.087    # physical drift (NOT used for pricing; Q uses r)
maturity <- 4        # maturity in years
M        <- 60       # number of time steps
sigma    <- 0.18     # volatility

K        <- 27000    # strike price of the European Call
coupon   <- 7.5      # fixed annual coupon of the variance certificate

n_paths  <- 10000    # simulated paths
set.seed(20260725)
```

---

# Simulating the DAX

Under $\mathbb{Q}$ the DAX follows
$$
dS_t = r\, S_t\, dt + \sigma\, S_t\, dW_t^{\mathbb{Q}},
$$
whose exact solution over a step of length $\Delta t$ is
$$
S_{t+\Delta t} = S_t \exp\!\left[\left(r - \tfrac{1}{2}\sigma^2\right)\Delta t
+ \sigma \sqrt{\Delta t}\, Z\right], \qquad Z \sim \mathcal{N}(0,1).
$$

`simulate_gbm()` implements this step exactly rather than by Euler
discretisation, so the terminal distribution is exact for any `M`. Only
path-dependent functionals are affected by the grid.

```{r simulate-paths}
paths <- simulate_gbm(S0, r, sigma, maturity, M, n_paths, antithetic = TRUE)
dim(paths)
```

## Sanity check

Under $\mathbb{Q}$ the expected terminal value must equal the forward price
$S_0 e^{r\tau}$. This is also asserted automatically in
`tests/testthat/test-gbm.R`; reproducing it here keeps the report readable on
its own.

```{r path-check}
data.frame(
  Quantity = c("Theoretical E[S_T] = S0 * exp(r * maturity)",
               "Simulated mean of S_T"),
  Value    = c(S0 * exp(r * maturity), mean(paths[, ncol(paths)]))
)
```

```{r path-plot, echo=FALSE}
matplot(t(paths[1:50, ]), type = "l", lty = 1,
        col = adjustcolor("steelblue", 0.5),
        xlab = "Time step (0 to 60)", ylab = "DAX level",
        main = "50 Simulated DAX Paths under the Risk-Neutral Measure")
abline(h = S0, col = "black", lwd = 2, lty = 2)
```

---

# European Call on the DAX

The call pays $\max(S_\tau - K, 0)$ and is **path-independent**, so only the
terminal value matters:
$$
C_0 = e^{-r\tau}\, \mathbb{E}^{\mathbb{Q}}\!\left[\max(S_\tau - K,\,0)\right].
$$

The payoff is monotone in the underlying shock, so antithetic sampling helps
here and the standard error is computed on the pairs.

```{r call-price}
call <- price_mc(payoff_call(paths, K), r, maturity, antithetic = TRUE)

data.frame(
  Price   = call$price,
  SE      = call$se,
  CI_low  = call$ci[1],
  CI_high = call$ci[2],
  ESS     = call$ess,
  n       = call$n
)
```

The effective sample size exceeds the number of paths: the antithetic
construction is buying genuine precision, which the pooled standard-error
formula used in the first version of this report concealed entirely.

## Validation against Black–Scholes

```{r call-bs}
bs <- bs_call(S0, K, r, sigma, maturity)

data.frame(
  Method     = c("Monte Carlo (antithetic)", "Black-Scholes closed form"),
  Price      = c(call$price, bs),
  Difference = c(call$price - bs, 0),
  In_CI      = c(NA, bs >= call$ci[1] && bs <= call$ci[2])
)
```

**Interpretation:** the analytic value lies inside the 95% confidence interval,
which validates the simulation engine end to end — path construction,
discounting and the standard error. The option is out-of-the-money at inception
(strike 27,000 against a spot of 25,250), so its value is mostly time value over
the four-year horizon.

---

# Variance Certificate

The certificate has two components:

1. **Fixed annual coupons** of 7.5 at the end of years 1 to 4.
2. A **terminal payoff** equal to the range of the DAX over the life of the
   certificate, divided by its average level:
$$
\frac{\max_{0\le t\le \tau} S_t - \min_{0\le t\le \tau} S_t}
     {\operatorname{avg}_{0\le t\le \tau} S_t}.
$$

$$
P_0 = \underbrace{\sum_{i=1}^{4} c\, e^{-r\,i}}_{\text{coupons}}
    \;+\; \underbrace{e^{-r\tau}\,\mathbb{E}^{\mathbb{Q}}
      \!\left[\frac{\max S_t - \min S_t}{\operatorname{avg} S_t}\right]}_{\text{terminal payoff}}.
$$

## Present value of the coupons

Deterministic, so no simulation is involved.

```{r certificate-coupons}
PV_coupons <- pv_coupons(coupon, r, maturity)

data.frame(
  Year       = seq_len(maturity),
  Coupon     = coupon,
  DiscFactor = exp(-r * seq_len(maturity)),
  PV         = coupon * exp(-r * seq_len(maturity))
)
```

## Terminal payoff

The range of a Brownian path is nearly invariant under a sign flip of the
shocks, so antithetic partners are strongly *positively* correlated here and the
technique works against us. The certificate is therefore priced on crude draws,
and the standard error is the ordinary one.

```{r certificate-terminal}
crude <- simulate_gbm(S0, r, sigma, maturity, M, n_paths, antithetic = FALSE)
term  <- price_mc(payoff_variance_certificate(crude), r, maturity,
                  antithetic = FALSE)

data.frame(Price = term$price, SE = term$se,
           CI_low = term$ci[1], CI_high = term$ci[2])
```

```{r certificate-hist, echo=FALSE}
pv <- payoff_variance_certificate(crude)
hist(pv, breaks = 40, col = "lightsteelblue", border = "white",
     xlab = "(max - min) / average",
     main = "Distribution of the Variance-Certificate Terminal Payoff")
abline(v = mean(pv), col = "red", lwd = 2)
```

## Arbitrage-free price

```{r certificate-price}
cert_price <- PV_coupons + term$price
cert_ci    <- cert_price + c(-1, 1) * 1.96 * term$se

data.frame(
  Component = c("PV of coupons", "PV of terminal payoff", "TOTAL"),
  Value     = c(PV_coupons, term$price, cert_price)
)
```

**Interpretation:** the price is dominated by the four fixed coupons. The
terminal payoff is a ratio of a range to an average and so is a small number,
well below one at this volatility. It rises close to linearly in $\sigma$ —
quantified in `04-sensitivity.Rmd`, where the slope is about 3.0 per unit of
volatility.

**Caveat on the grid.** The maximum and minimum are taken over 61 observation
points, while the contract specifies continuous monitoring. The discrete range
can only understate the continuous one, so the terminal payoff above is biased
downwards — by roughly 10% at this grid, which is around 25 times the Monte
Carlo standard error. The confidence interval quoted here describes simulation
noise only and says nothing about that bias. See `03-discretisation.Rmd` for the
convergence study and an extrapolated continuous-monitoring value.

---

# Conclusion

```{r results-summary, echo=FALSE}
data.frame(
  Instrument = c("European Call (K = 27000)", "Variance Certificate (coupon = 7.5)"),
  Price      = c(round(call$price, 2), round(cert_price, 4)),
  SE         = c(round(call$se, 2), round(term$se, 4)),
  Sampling   = c("antithetic", "crude")
)
```

Both instruments were priced under the risk-neutral measure on 10,000 paths:

* The **European Call** was valued from the discounted terminal payoff and
  agrees with the Black–Scholes closed form to within Monte Carlo error.
* The **Variance Certificate** was valued as the discounted coupon stream plus
  the expected terminal ratio, using the full simulated paths.

The modelling choices — riskless drift rather than physical, exact GBM stepping,
full paths for the path-dependent leg — follow from no-arbitrage pricing theory.
The choices that are *not* standard, and that the companion reports justify,
are using antithetic sampling for only one of the two instruments and computing
the standard error on pairs.

## Limitations

* The certificate value is a lower bound: the 60-step grid understates the
  continuous range by roughly 10%.
* Volatility is assumed, not calibrated. It is the largest single source of
  error in the price — larger than the grid bias and far larger than Monte Carlo
  noise.
* Both prices are model prices under geometric Brownian motion, with constant
  volatility and no jumps, and are not calibrated to traded DAX option prices.
