---
title: "Does Antithetic Sampling Actually Help?"
author: "Daniil Rabotnev"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: paper
    highlight: tango
    df_print: paged
subtitle: Measuring the variance reduction instead of assuming it
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 5,
                      cache = TRUE)
devtools::load_all("..")
```

# The question

Antithetic variates are the standard first move in Monte Carlo pricing: draw
$Z$, also use $-Z$, get a lower variance for free. The pricing report applies
them to both instruments and reports a confidence interval based on
`sd(payoffs) / sqrt(n)`.

Two things are worth checking rather than assuming.

1. That standard error treats the $n$ payoffs as independent. They are not —
   antithetic sampling produces $n/2$ *pairs*. Is the reported interval right?
2. Variance reduction is not automatic. It relies on the payoff being monotone
   in the underlying shock. Does that hold for both instruments here?

This report answers both by measurement.

```{r parameters}
S0       <- 25250
r        <- 0.02
sigma    <- 0.18
maturity <- 4
M        <- 60
K        <- 27000

n_paths  <- 10000
n_reps   <- 200      # replications for the ground-truth study
```

# A single run

```{r single-run}
set.seed(20260725)
paths <- simulate_gbm(S0, r, sigma, maturity, M, n_paths, antithetic = TRUE)

payoffs <- list(
  call        = payoff_call(paths, K),
  certificate = payoff_variance_certificate(paths)
)

diagnose <- function(p) {
  half   <- length(p) / 2
  pooled <- price_mc(p, r, maturity, antithetic = FALSE)
  paired <- price_mc(p, r, maturity, antithetic = TRUE)
  data.frame(
    price       = paired$price,
    pooled_se   = pooled$se,
    paired_se   = paired$se,
    correlation = stats::cor(p[seq_len(half)], p[half + seq_len(half)]),
    ess         = paired$ess
  )
}

single <- do.call(rbind, lapply(payoffs, diagnose))
single
```

The two rows already disagree in sign. For the call the antithetic partners are
negatively correlated, which is what the technique is supposed to deliver. For
the certificate the correlation is close to $+1$ — the partner path carries
almost the same information as the original.

The effective sample size makes the consequence concrete: the call behaves as
if it had far more than `r format(n_paths, big.mark = ",")` independent draws,
the certificate as if it had roughly half.

# Which standard error is right?

A standard error is a claim about how much the price estimate would move if the
simulation were rerun. That claim is directly testable: rerun it many times and
look at the spread.

```{r replication-study}
one_replication <- function(seed) {
  set.seed(seed)
  P <- simulate_gbm(S0, r, sigma, maturity, M, n_paths, antithetic = TRUE)

  rows <- lapply(
    list(call        = payoff_call(P, K),
         certificate = payoff_variance_certificate(P)),
    function(p) {
      data.frame(
        price     = price_mc(p, r, maturity, antithetic = TRUE)$price,
        pooled_se = price_mc(p, r, maturity, antithetic = FALSE)$se,
        paired_se = price_mc(p, r, maturity, antithetic = TRUE)$se
      )
    }
  )
  cbind(instrument = names(rows), do.call(rbind, rows))
}

runs <- do.call(rbind, lapply(seq_len(n_reps), one_replication))

comparison <- do.call(rbind, lapply(split(runs, runs$instrument), function(d) {
  truth <- stats::sd(d$price)
  data.frame(
    empirical_sd  = truth,
    pooled_se     = mean(d$pooled_se),
    pooled_ratio  = mean(d$pooled_se) / truth,
    paired_se     = mean(d$paired_se),
    paired_ratio  = mean(d$paired_se) / truth
  )
}))
comparison
```

`empirical_sd` is the ground truth: the actual spread of the price across
`r n_reps` independent reruns. The two `*_ratio` columns say how badly each
formula misstates it.

The pooled standard error is wrong in both cases, and wrong in opposite
directions — too wide for the call, too narrow for the certificate. The paired
standard error tracks the truth for both.

```{r se-plot, echo=FALSE, fig.height=4}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
for (nm in rownames(comparison)) {
  d <- comparison[nm, ]
  bp <- barplot(c(Truth = d$empirical_sd, Pooled = d$pooled_se,
                  Paired = d$paired_se),
                col = c("grey40", "indianred", "steelblue"),
                ylab = "standard error", main = nm)
  abline(h = d$empirical_sd, lty = 2)
}
par(op)
```

# Why the certificate breaks it

Antithetic sampling reduces variance when the payoff is monotone in the shock,
because then $f(Z)$ and $f(-Z)$ move in opposite directions. The certificate
payoff is not monotone — it is a *range*.

For a driftless Brownian path the range is exactly invariant under a sign flip:

$$
\max_t(-W_t) - \min_t(-W_t) = -\min_t W_t + \max_t W_t = \max_t W_t - \min_t W_t .
$$

That identity is exact in *log* space. The payoff, however, is measured in
price space, and $\exp(\cdot)$ is monotone but not linear: it preserves where
the maximum and minimum occur but not the distance between them. That is the
only thing standing between the measured correlation and a perfect $+1$.

The demonstration below separates the two. Choosing $r - \sigma^2/2 = 0$ kills
the drift, so the log-range columns must agree to machine precision, while the
price-space payoffs merely come close.

```{r reflection-demo}
set.seed(7)
flat <- simulate_gbm(S0 = 100, r = 0.02, sigma = 0.2,   # r - sigma^2/2 == 0
                     maturity = 1, M = 250, n_paths = 6, antithetic = TRUE)

log_range <- function(P) apply(log(P), 1, function(x) max(x) - min(x))

data.frame(
  log_range_original   = log_range(flat)[1:3],
  log_range_antithetic = log_range(flat)[4:6],
  payoff_original      = payoff_variance_certificate(flat)[1:3],
  payoff_antithetic    = payoff_variance_certificate(flat)[4:6]
)
```

The first two columns are identical, the last two are not. The antithetic
partner is close to a duplicate, so half the simulation budget buys almost
nothing.

# Crude versus antithetic at equal cost

The fair comparison holds the number of simulated paths fixed and asks which
scheme gives the smaller true error.

```{r crude-vs-antithetic}
scheme_sd <- function(antithetic, seed_offset) {
  prices <- vapply(seq_len(n_reps), function(i) {
    set.seed(seed_offset + i)
    P <- simulate_gbm(S0, r, sigma, maturity, M, n_paths, antithetic = antithetic)
    c(price_mc(payoff_call(P, K), r, maturity)$price,
      price_mc(payoff_variance_certificate(P), r, maturity)$price)
  }, numeric(2))
  stats::setNames(apply(prices, 1, stats::sd), c("call", "certificate"))
}

head_to_head <- rbind(
  crude      = scheme_sd(FALSE, 90000),
  antithetic = scheme_sd(TRUE,  95000)
)
rbind(head_to_head, ratio = head_to_head["antithetic", ] / head_to_head["crude", ])
```

A ratio below one means antithetic sampling wins. Below the call column it does;
below the certificate column it does not.

# Conclusion

* The standard error must be computed on the $n/2$ pair averages, not on the
  pooled $n$ payoffs. Doing it the pooled way misstated the precision by roughly
  a quarter in both instruments, in opposite directions.
* Antithetic sampling is worth using for the European call and actively
  counterproductive for the variance certificate, because the range of a path is
  near-invariant under the sign flip the technique relies on.
* Reporting the effective sample size alongside the standard error makes this
  visible at a glance: values below $n$ are a signal that the variance-reduction
  scheme is not paying for itself.

The practical recommendation for this pricing problem is to keep antithetic
sampling for the call and price the certificate with crude Monte Carlo at the
same path count.
