---
title: "How Much Does the Time Grid Cost?"
author: "Daniil Rabotnev"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: paper
    highlight: tango
    df_print: paged
subtitle: Discretisation bias in a path-dependent payoff
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 5,
                      cache = TRUE)
devtools::load_all("..")
```

# The question

The certificate pays

$$
\frac{\max_{0 \le t \le \tau} S_t - \min_{0 \le t \le \tau} S_t}
     {\operatorname{avg}_{0 \le t \le \tau} S_t},
$$

where the maximum and minimum are taken over *continuous* time. The simulation
observes the path at 61 points and takes the maximum and minimum of those.

Between two grid points the path keeps moving. Whatever it does there is
invisible, and it can only push the observed maximum down and the observed
minimum up relative to the truth. The discrete range is therefore **always** at
most the continuous range, never more. This is a one-sided error — a bias, not
noise — and adding paths does not remove it.

The previous report established how to measure Monte Carlo error correctly.
This one asks whether that error is the thing worth worrying about.

```{r parameters}
S0       <- 25250
r        <- 0.02
sigma    <- 0.18
maturity <- 4
K        <- 27000

M_grid   <- c(60, 125, 250, 500, 1000)
n_paths  <- 20000
```

Antithetic sampling is switched off here. Report 02 showed it does nothing for
this payoff, and leaving it on would only make the standard errors harder to
read.

# Refining the grid

```{r convergence}
grid_run <- function(M) {
  set.seed(4242)
  P    <- simulate_gbm(S0, r, sigma, maturity, M, n_paths, antithetic = FALSE)
  cert <- price_mc(payoff_variance_certificate(P), r, maturity)
  call <- price_mc(payoff_call(P, K), r, maturity)
  data.frame(M = M,
             certificate = cert$price, certificate_se = cert$se,
             call = call$price, call_se = call$se)
}

grid <- do.call(rbind, lapply(M_grid, grid_run))
grid
```

The certificate price rises monotonically with the number of steps, exactly as
the argument above predicts. Every refinement uncovers a little more of the
path's true excursion.

# Extrapolating to continuous monitoring

For discretely monitored path-dependent claims the bias decays like
$1/\sqrt{M}$ rather than $1/M$ — the slow rate is a known feature of maxima of
Brownian paths, not an artefact of this implementation. That gives a way to
estimate the continuous-monitoring price without ever simulating it: fit

$$
P(M) \;=\; a + \frac{b}{\sqrt{M}}
$$

and read off the intercept $a$, which is the limit as $M \to \infty$.

```{r extrapolation}
fit <- stats::lm(certificate ~ I(1 / sqrt(M)), data = grid)

continuous_limit <- unname(stats::coef(fit)[1])
bias_at_60       <- grid$certificate[grid$M == 60] - continuous_limit

data.frame(
  intercept_a      = continuous_limit,
  slope_b          = unname(stats::coef(fit)[2]),
  r_squared        = summary(fit)$r.squared,
  price_at_M_60    = grid$certificate[grid$M == 60],
  bias_at_M_60     = bias_at_60,
  bias_percent     = 100 * bias_at_60 / continuous_limit
)
```

```{r convergence-plot, echo=FALSE}
xs <- 1 / sqrt(grid$M)
plot(xs, grid$certificate, pch = 19, col = "steelblue",
     xlim = c(0, max(xs) * 1.05),
     ylim = range(c(grid$certificate, continuous_limit)) + c(-0.01, 0.01),
     xlab = expression(1 / sqrt(M)), ylab = "certificate price",
     main = "Convergence to continuous monitoring")
abline(fit, col = "grey40")
points(0, continuous_limit, pch = 4, cex = 1.6, lwd = 2, col = "indianred")
text(xs, grid$certificate, labels = paste0("M=", grid$M), pos = 4, cex = 0.8)
text(0, continuous_limit, labels = "extrapolated limit", pos = 4,
     cex = 0.8, col = "indianred")
```

Plotting against $1/\sqrt{M}$ turns the convergence into a straight line, and
the intercept at zero is the continuous-monitoring price. The fit being close
to linear is itself evidence that the $1/\sqrt{M}$ rate is the right one.

# The control: a payoff that cannot be affected

If the simulator itself were mis-specified, everything above would be
meaningless. The European call provides a clean control. It depends only on
$S_\tau$, and the GBM step is exact rather than an Euler approximation, so its
price must not depend on $M$ at all.

```{r control}
data.frame(
  M          = grid$M,
  call       = grid$call,
  call_se    = grid$call_se,
  z_vs_first = (grid$call - grid$call[1]) / grid$call_se
)
```

The call price wanders by well under two standard errors across a
seventeen-fold refinement of the grid — pure Monte Carlo noise, no trend. The
simulator is fine. The bias in the certificate comes from the payoff being
path-dependent, nothing else.

# Bias against noise

```{r error-comparison}
mc_se <- grid$certificate_se[grid$M == 60]

data.frame(
  source = c("Monte Carlo standard error", "Discretisation bias"),
  size   = c(mc_se, abs(bias_at_60)),
  ratio  = c(1, abs(bias_at_60) / mc_se)
)
```

This is the finding. At `r sprintf("M = %d", 60)` the reported confidence
interval is roughly `r sprintf("%.4f", 2 * 1.96 * mc_se)` wide, while the
number it brackets sits about `r sprintf("%.4f", abs(bias_at_60))` away from
the quantity the contract actually specifies. The interval is precise and
centred on the wrong value.

Increasing the number of paths narrows the interval and does nothing about the
gap. Only refining the grid — or correcting for it, as the extrapolation above
does — moves the estimate towards the truth.

# Conclusion

* The discrete maximum and minimum bias the certificate payoff downwards by
  construction. At `M = 60` the price is understated by roughly
  `r sprintf("%.0f%%", abs(100 * bias_at_60 / continuous_limit))`.
* The bias decays like $1/\sqrt{M}$, so a linear fit against $1/\sqrt{M}$
  recovers the continuous-monitoring price without simulating it.
* At the grid used in the original report the bias is around
  `r sprintf("%.0f", abs(bias_at_60) / mc_se)` times the Monte Carlo standard
  error. Reporting a tight confidence interval at that resolution overstates
  what the simulation knows.
* The European call is unaffected, which confirms the simulator is correct and
  isolates the cause to path dependence.

The practical recommendation is to price the certificate on a grid of at least
several hundred steps, or to quote the extrapolated value with the fit above
and state the grid it came from.
