---
title: "Which Error Actually Matters?"
author: "Daniil Rabotnev"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: paper
    highlight: tango
    df_print: paged
subtitle: Volatility sensitivity and an error budget for the certificate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 5,
                      cache = TRUE)
devtools::load_all("..")
```

# The question

The pricing report claims the certificate payoff rises with volatility, and
leaves it at that. Two things follow from taking the claim seriously.

First, how strongly? A payoff defined as a range divided by an average is not
obviously linear in $\sigma$, and the shape matters for hedging intuition.

Second, and more usefully: volatility is not observed. It is estimated, and the
estimate is wrong by some amount. That is a third source of error alongside the
two already measured — Monte Carlo noise in report 02, grid bias in report 03.
Ranking all three says where effort should go.

```{r parameters}
S0       <- 25250
r        <- 0.02
sigma    <- 0.18
maturity <- 4
K        <- 27000

M        <- 250        # fine enough that grid bias is small, per report 03
n_paths  <- 40000
sigma_grid <- c(0.05, 0.10, 0.15, 0.18, 0.25, 0.35, 0.50)
```

# Sweeping volatility

```{r sweep}
price_at <- function(s, M. = M, n. = n_paths, seed = 808) {
  set.seed(seed)
  P    <- simulate_gbm(S0, r, s, maturity, M., n., antithetic = FALSE)
  cert <- price_mc(payoff_variance_certificate(P), r, maturity)
  call <- price_mc(payoff_call(P, K), r, maturity)
  data.frame(sigma = s,
             certificate = cert$price, certificate_se = cert$se,
             call = call$price, call_se = call$se)
}

sweep <- do.call(rbind, lapply(sigma_grid, price_at))
sweep
```

```{r sweep-plot, echo=FALSE, fig.height=4}
op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

with(sweep, {
  plot(sigma, certificate, type = "b", pch = 19, col = "steelblue",
       xlab = expression(sigma), ylab = "price",
       main = "Variance certificate")
  arrows(sigma, certificate - 1.96 * certificate_se,
         sigma, certificate + 1.96 * certificate_se,
         angle = 90, code = 3, length = 0.03, col = "grey50")
  abline(lm(certificate ~ sigma), lty = 2, col = "indianred")

  plot(sigma, call, type = "b", pch = 19, col = "steelblue",
       xlab = expression(sigma), ylab = "price",
       main = "European call")
  arrows(sigma, call - 1.96 * call_se, sigma, call + 1.96 * call_se,
         angle = 90, code = 3, length = 0.03, col = "grey50")
})
par(op)
```

The confidence intervals are too narrow to see at this scale, which is the
first hint at the answer to the second question.

# The certificate is linear in volatility

```{r linearity}
lin <- stats::lm(certificate ~ sigma, data = sweep)

data.frame(
  intercept = unname(stats::coef(lin)[1]),
  slope     = unname(stats::coef(lin)[2]),
  r_squared = summary(lin)$r.squared,
  max_abs_residual = max(abs(stats::residuals(lin)))
)
```

A dimensionless range-over-average payoff turns out to be almost exactly
proportional to volatility over the whole range from 5% to 50%. The slope is
the useful number: it converts a volatility error directly into a price error.

The call, by contrast, is visibly concave in $\sigma$ — the familiar shape of
vega falling away as the option moves deep in the money.

# An error budget

Three sources of error, measured in the units of the certificate price at
$\sigma = 18\%$.

```{r error-budget}
base <- price_at(sigma)

# 1. Monte Carlo noise, at the settings used in the original pricing report
set.seed(20260725)
P_orig <- simulate_gbm(S0, r, sigma, maturity, M = 60, n_paths = 10000,
                       antithetic = TRUE)
mc_se <- price_mc(payoff_variance_certificate(P_orig), r, maturity,
                  antithetic = TRUE)$se

# 2. Grid bias: how much the price moves from M = 60 to a fine grid.
#    Report 03 extrapolates properly; this is a conservative lower bound.
coarse    <- price_at(sigma, M. = 60)$certificate
fine      <- price_at(sigma, M. = 1000)$certificate
grid_bias <- fine - coarse

# 3. Volatility misspecification: three vol points either way
vol_error <- unname(stats::coef(lin)[2]) * 0.03

budget <- data.frame(
  source = c("Monte Carlo noise (n = 10,000)",
             "Grid bias (M = 60 vs M = 1000)",
             "Volatility off by 3 points"),
  size   = c(mc_se, abs(grid_bias), abs(vol_error))
)
budget$relative_to_noise <- budget$size / budget$size[1]
budget
```

```{r budget-plot, echo=FALSE, fig.height=3.5}
op <- par(mar = c(4, 14, 2, 1))
barplot(rev(budget$size), horiz = TRUE, las = 1,
        names.arg = rev(budget$source), col = "steelblue",
        xlab = "error in certificate price units",
        main = "Where the uncertainty actually comes from")
par(op)
```

# Conclusion

* The certificate price is linear in volatility with a slope of about
  `r sprintf("%.1f", unname(stats::coef(lin)[2]))` per unit of $\sigma$, so
  each volatility point is worth roughly
  `r sprintf("%.3f", unname(stats::coef(lin)[2]) / 100)` in price.
* Monte Carlo noise is the *smallest* of the three error sources by more than
  an order of magnitude. Running more paths is the least useful thing that
  could be done to this model.
* The grid bias is larger, and cheap to fix: refine the grid or extrapolate.
* Volatility misspecification dominates everything. Nothing inside the
  simulation can address it.

The ordering points outside the model. The next useful step is not a numerical
improvement but an empirical one: estimate $\sigma$ from realised DAX returns
and from option-implied volatility, and price the certificate on both, so the
quoted number carries a market-based input rather than an assumed one.
